---
title: "BiocHubQuery"
author: "BiocHubs Team"
format: html
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{BiocHubQuery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, vignette_setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

# Background

AnnotationHub and ExperimentHub were developed within Bioconductor to facilitate
the distribution of large annotation and experimental datasets without requiring
those resources to be included directly in software packages. This approach
helps avoid package bloat and keeps package sizes manageable while still
providing access to substantial external resources.

Traditionally, using the AnnotationHub or ExperimentHub packages requires
downloading a SQLite database containing metadata describing all available
resources before those resources can be searched or queried. As the number of
available resources continues to grow, the size of this metadata database
increases accordingly.

BiocHubQuery provides an alternative approach. A separate data pipeline converts
the AnnotationHub and ExperimentHub SQLite metadata tables into Parquet files
that are hosted remotely. BiocHubQuery connects directly to these Parquet files,
allowing users to query and filter metadata without first downloading the full
SQLite database.

# Installation and Loading

`BiocHubQuery` is a _Bioconductor_ package and can be installed through
`BiocManager::install()`.

```{r, install, eval = FALSE}
if (!"BiocManager" %in% rownames(installed.packages()))
     install.packages("BiocManager")
BiocManager::install("BiocHubQuery", dependencies=TRUE)
```

After the package is installed, it can be loaded into _R_ workspace by

```{r, library, results='hide', warning=FALSE, message=FALSE}
library(BiocHubQuery)
```

We will also load libraries need to run this vignette:

```{r, setup}
library(BiocHubQuery)
library(dplyr)
```

# Available Tables

The AnnotationHub and ExperimentHub databases share a common SQLite schema. In
`BiocHubQuery`, each table in this schema has been converted into a Parquet file
and made available for remote querying. In addition to the raw tables, a set of
pre-computed views is also provided to simplify common join operations.

## Existing SQLite Tables

The following tables originate directly from the SQLite database and are
available as Parquet files: resources, rdatapaths, location_prefixes,
input_sources, statuses, and tags.

#### resources

The central table describing all available Hub resources. Each record
corresponds to a single dataset or resource registered in AnnotationHub or
ExperimentHub.

Key fields include metadata such as:

  - resource title and description
  - species and genome information
  - data provider and maintainer
  - status and lifecycle information
  - links to related tables via foreign keys

#### rdatapaths

Defines one or more R data object representations/files associated with each
resource.

This table includes:

  - file paths for data objects
  - R data classes (e.g., GRanges, SummarizedExperiment)
  - dispatch information for object retrieval

This table is typically used to determine how a resource should be loaded into
R and combined with information in location_prefixes gives a complete download
url of a file.

**Note:** A single resource in the `resources` table may be associated with
  multiple entries in the `rdatapaths` table. For example, a BAM resource will
  have one entry in `resources`, while its associated files (e.g., BAM and BAI)
  are represented as separate entries in `rdatapaths`. The id of the `resources`
  table will map to the resource_id of the `rdatapaths` table.
  

#### location_prefixes

Provides location metadata used to construct or interpret resource storage
paths.

This is primarily used internally to resolve where a resource is located within
a remote or local storage system.

The `resources` table location_prefix_id maps to an id of this table.


#### input_sources

Contains provenance information describing the original source of each resource.

This includes:

  - source URLs
  - version information
  - checksums (MD5)
  - last modified dates
  - file sizes and source types

This table is useful for tracking data origin. The id of the `resources` table
will map to the resource_id of this table.

#### statuses

Defines lifecycle status information for each resource. Typical values indicate
whether a resource is active, deprecated, or removed. The `resources` table
status_id maps to the id of this table.

#### tags

Associates free-form tags with resources to support categorisation and search.
Each tag is linked to a resource_id that maps back to the `resources` table id.

#### Other (non exposed)

The underlying SQLite schema also contains additional tables used for internal
bookkeeping: biocversions, recipes, schema_info, and timestamp. These tables are
not currently exposed through BiocHubQuery, as they primarily support internal
versioning and infrastructure concerns rather than user-facing queries.

If access to these tables becomes necessary for advanced use cases, they may be
exposed upon request.

## Recommended Additional Views 

To simplify common analytical workflows, BiocHubQuery also provides pre-computed
views. These are derived from joins across multiple tables and are generally
recommended over working with the above listed raw tables. They include:
resource_metadata, resource_downloadlinks, and resource_full.


#### resource_metadata

A denormalized view combining core resources table information with:

  - status labels
  - location prefix metadata
  - aggregated R data class information

This view provides a convenient “one-row-per-resource” summary.

#### resource_downloadlinks

Provides resolved download paths for each resource.

This includes:

  - resource identifiers
  - resolved file paths
  - status information
  - constructed full storage paths

Useful for programmatic download workflows.

#### resource_full

A comprehensive joined view combining metadata and download information.

This view is intended as the primary entry point for most users, providing:

  - full resource metadata
  - file-level download information
  - associated R data class mappings

It is designed to support efficient filtering and querying without requiring
manual joins across multiple tables.

# Retrieving Data

Example query

# SessionInfo

```{r, sessioninfo}
sessionInfo()
```